---
title: "Assignment 2"
author: "Colin Wick"
date: "2/10/2021"
output: github_document
always_allow_html: true
---

```{r,include=FALSE}
knitr::opts_chunk$set(echo = FALSE,error = FALSE,warning = FALSE,message = FALSE)
```

```{r,include=FALSE}
library(tidyverse)
library("rddensity")
library(lubridate)
library("kableExtra")
library("rdd")
```

```{r,cache=TRUE}
mydf <- read.csv("https://raw.githubusercontent.com/scunning1975/causal-inference-class/master/Data/hansen_dwi.csv")

mydf$day <- substr(mydf$Date,1,2)
mydf$month <- as.factor(substr(mydf$Date,3,5))
levels(mydf$month) <- c(04,08,12,02,01,07,06,03,05,11,10,09)

mydf$Date <- as.Date(paste(mydf$day,mydf$month,mydf$year,sep = "-"),format = "%d-%m-%Y")
```

Hansen (2015) uses panel BAC and police data to construct a regression discontinuity model exploiting the "arbitrary" cutoffs of DUI consequences at different BAC levels. The first task is establishing the homogeneity and independence of BAC in order to validate that there is no clustering around specific values, as shown in Figure 1. The BAC of drivers is approximately normal with no evidence of bunching around cutoffs.

Then, exploiting the underlying normality of BAC levels, he is able to analyze differences between those who fell just above and just below a given threshold and DUI recidivism. He finds a sharp drop in recidivism along the first BAC cutoff (.08) and a weaker drop on the second.

He then does a handful of robustness checks and deeper looks into recidivism, finding the effect is strongest for those whom had never been tested and were caught above the .08 threshold but below the .15 threshold. 

```{r}
mydf <- mydf %>%
  mutate(bac_08 = ifelse(bac1 >= .08,1,0),
         bac_15 = ifelse(bac1 >= .15,1,0))

mydf %>%
  ggplot()+
  geom_histogram(aes(bac1),binwidth = .003,fill="light grey",color="grey")+
  geom_vline(xintercept = c(.08,.15),size=1)+
  theme_minimal()

rdd_test08 <- rddensity(X = mydf$bac1,c=.08)
rdd_test15 <- rddensity(X = mydf$bac1,c=.15)
print(paste("with P-value of McCrary tests",round(rdd_test08$test$p_jk,2),"and",round(rdd_test15$test$p_jk,2),"we fail to reject the null that there is no bunching around cutoffs"))
```

Recreating Figure 1 from Hansen. A histogram of BAC shows there is no visible evidence of bunching around either cutoff. Running the McCrary test finds no statistical evidence either.

```{r}
table_2_rep_08 <- data.frame()

rdd_08m <- mydf %>%
  #filter((bac1 <= .085 & mydf$bac1 >= .075) | bac1 <= .145 & mydf$bac1 >= .155) %>%
  rdd::RDestimate(formula = male ~ bac1 + bac_08 + bac1 * bac_08,cutpoint = .08,bw = .05,kernel = "rectangular")
rdd_08a <- mydf %>%
  #filter((bac1 <= .085 & mydf$bac1 >= .075) | bac1 <= .145 & mydf$bac1 >= .155) %>%
  rdd::RDestimate(formula = aged ~ bac1 + bac_08 + bac1 * bac_08,cutpoint = .08,bw = .05,kernel = "rectangular")
rdd_08w <- mydf %>%
  #filter((bac1 <= .085 & mydf$bac1 >= .075) | bac1 <= .145 & mydf$bac1 >= .155) %>%
  rdd::RDestimate(formula = white ~ bac1 + bac_08 + bac1 * bac_08,cutpoint = .08,bw = .05,kernel = "rectangular")

table_2_rep_08[1,1] <- rdd_08m$est[1]
table_2_rep_08[2,1] <- rdd_08m$se[1]
table_2_rep_08[1,2] <- rdd_08w$est[1]
table_2_rep_08[2,2] <- rdd_08w$se[1]
table_2_rep_08[1,3] <- rdd_08a$est[1]
table_2_rep_08[2,3] <- rdd_08a$se[1]

table_2_rep_08[3,] <- mydf %>%
  filter(bac1 > .029 & bac1 < .129) %>%
  summarize(malemean = mean(male),
            whitemean = mean(white),
            agemean = mean(aged))

names(table_2_rep_08) <- c("male","white","age")
row.names(table_2_rep_08) <- c("estimate","se","mean at .079")
```

```{r}
table_2_rep_15 <- data.frame()

rdd_15m <- mydf %>%
  #filter((bac1 <= .155 & mydf$bac1 >= .075) | bac1 <= .145 & mydf$bac1 >= .155) %>%
  rdd::RDestimate(formula = male ~ bac1 + bac_15 + bac1 * bac_15,cutpoint = .15,bw = .05,kernel = "rectangular")
rdd_15a <- mydf %>%
  #filter((bac1 <= .155 & mydf$bac1 >= .075) | bac1 <= .145 & mydf$bac1 >= .155) %>%
  rdd::RDestimate(formula = aged ~ bac1 + bac_15 + bac1 * bac_15,cutpoint = .15,bw = .05,kernel = "rectangular")
rdd_15w <- mydf %>%
  #filter(bac1 <= .9 & mydf$bac1 >= .14) %>%
  rdd::RDestimate(formula = white ~ bac1 + bac_15 + bac1 * bac_15,cutpoint = .15,bw = .05,kernel = "rectangular")

table_2_rep_15[1,1] <- rdd_15m$est[1]
table_2_rep_15[2,1] <- rdd_15m$se[1]
table_2_rep_15[1,2] <- rdd_15w$est[1]
table_2_rep_15[2,2] <- rdd_15w$se[1]
table_2_rep_15[1,3] <- rdd_15a$est[1]
table_2_rep_15[2,3] <- rdd_15a$se[1]

table_2_rep_15[3,] <- mydf %>%
  filter(bac1 > .099 & bac1 < .199) %>%
  summarize(malemean = mean(male),
            whitemean = mean(white),
            agemean = mean(aged))
names(table_2_rep_15) <- c("male","white","age")
row.names(table_2_rep_15) <- c("estimate","se","mean at .149")

knitr::kables(
  list(kbl(round(table_2_rep_08,c(4,3,2)),caption = "For BAC=.08"),
       kbl(round(table_2_rep_15,c(4,3,2)),caption = "For BAC=.15"))) %>%
  kable_classic_2()
```

```{r,fig.align='center',fig.width=9,fig.height=6}
library(splines)
library(matrixStats)


mydf %>%
  mutate(bac_bin = cut(x = mydf$bac1,breaks=seq(0,max(mydf$bac1),.002),right = FALSE)) %>%
  filter(bac1 <= .2) %>%
  group_by(bac_bin) %>%
  summarize(acc_avg = mean(acc),
            male_avg = mean(male),
            age_avg = mean(aged),
            wht_avg = mean(white),
            bac_mid = max(bac1)) %>%
  pivot_longer(c(acc_avg,male_avg,age_avg,wht_avg)) %>%
  #mutate(backnot = ifelse(bac_mid < .08,".08",ifelse(bac_mid >= .08 & bac_mid < .15,".15","over"))) %>%
  mutate(backnot = ifelse(bac_mid > .08,1,0)) %>%
  ggplot()+
  geom_point(aes(bac_mid,value),alpha=.25)+
  stat_smooth(aes(bac_mid,value,group=backnot),
              method = lm,
              se = TRUE,size=1)+
  facet_wrap(~ name, scales = 'free', nrow = 2)
```

```{r}
head(mydf$bac2)

mydf$sq_bac1 <- mydf$bac1^2

f1 <- formula(recidivism ~ bac1 | white + aged + male)
f2 <- formula(recidivism ~ bac1 + bac_08 + bac1 * bac_08 | white + aged + male)
f3 <- formula(recidivism ~ bac1 + bac_08 + sq_bac1 | white + aged + male)


estimates_.05_1 <- rdd::RDestimate(data=mydf,
                                   formula = f1,
                                   cutpoint = .08,
                                   kernel = "rectangular",
                                   bw = .05)

estimates_.05_2 <- rdd::RDestimate(data=mydf,
                                   formula = f2,
                                   cutpoint = .08,
                                   kernel = "rectangular",
                                   bw = .05)

estimates_.05_3 <- rdd::RDestimate(data=mydf,
                                   formula = f3,
                                   cutpoint = .08,
                                   kernel = "rectangular",
                                   bw = .05)


table_3 <- data.frame()
estimates_.05_1$est
estimates_.05_2$est
estimates_.05_3$est
table_3[c(1:2),1] <- c(estimates_.05_1$est[1],estimates_.05_1$se[1])
table_3[c(1:2),2] <- c(estimates_.05_2$est[1],estimates_.05_2$se[1])
table_3[c(1:2),3] <- c(estimates_.05_3$est[1],estimates_.05_3$se[1])
table_3
```
